Title Goes Here

### 2023-01-04 inst/paper/_intro.qmd
add: revisions based on co-authors feedback

<!-- intro to calculus and residues -->
Dental calculus has proven to be excellent source of a wide variety of information
about our past.

the increased availability of methods such as aDNA, paleoproteomics, and
chromatography, has expanded our ability to identify specific pathogens, dietary,
and possible medicinal compounds<!--cite-->.

<!--how are compounds incorporated in dental calculus-->
Increased interest in using oral fluid<!--more accurate terminology than saliva?--> as a means of detecting alkaloids in living
individuals, due to the non-invasive nature of the testing compared to blood
and urine sampling<!--citations-->.

Alkaloids enter the dental calculus via two routes: (1) direct
incorporation through oral consumption of alkaloid-containing plants, whether
deliberate or accidental;
and (2) passive diffusion through the
saliva as toxins are transferred from serum to saliva [@coneInterpretationOral2007],
and then into the oral cavity through the salivary glands. The relation to serum
is why there is often a close relation between oral fluid and blood concentrations
of drugs [@coneInterpretationOral2007; @sorensenDrugsCalculus2021]
The second route allows the
detection of compounds that not only are consumed orally, but also any other pathway.
We also expect to detect metabolites through the second pathway.

Many of the components involved in the formation and growth of dental calculus
originate from saliva. Therefore the transfer of salts and other
compounds between saliva and oral biofilms on the tooth surface is likely to also
allow various alkaloids of dietary and medicinal origin to become entrapped in
dental plaque, eventually mineralising and becoming dental calculus.

Previous studies have been conducted using various chromatographic methods on
archaeological dental calculus, employing both targeted
[@eerkensDentalCalculus2018] and untargeted approaches
[@buckleyDentalCalculus2014; @gismondiMultidisciplinaryApproach2020].

<!--brief description of method validation-->
In the method validation study, all compounds detected in the blood were also
detected in the dental calculus,
with additional compounds present in dental calculus that were not present
in blood, suggesting that dental calculus represents a comprehensive history of
consumption over a long time <!-- rephrase-->[@sorensenDrugsCalculus2021].

<!--summary of results-->
We were able to detect both parent compounds, such as caffeine and nicotine, as
well as their metabolites theophylline and cotinine. These indicate the consumption
of tea and coffee and smoking of tobacco, which is also confirmed by historic
documentation and identification of pipe notches in the dentition.

>>>>

<!-- intro to calculus and residues -->
Dental calculus has proven to be an excellent source of a wide variety of information
about our past. The increased accessibility and advancement of methods such as
aDNA, paleoproteomics, and chromatography, has expanded our ability to identify
biomarkers of diet and disease on an increasingly large scale
[@warinnerEvidenceMilk2014; @velskoDentalCalculus2017; @gismondiMultidisciplinaryApproach2020].

<!-- define alkaloids -->
One such biomarker is alkaloids, a plant-derived group of compounds. Many alkaloids
have important medicinal and psychoactive effects in humans, and their direct
detection, or detection of their metabolites, is of great interest to archaeologists.
Previous studies have successfully recovered alkaloids archaeological
contexts including ceramics [@smithDetectionOpium2018], pipes
[@raffertyCurrentResearch2012], human hair
[@ogaldeIdentificationPsychoactive2009; @echeverriaNicotineHair2013], and even
dental calculus employing both targeted [@eerkensDentalCalculus2018] and
untargeted approaches
[@buckleyDentalCalculus2014; @gismondiMultidisciplinaryApproach2020].
Especially nicotine, the principal alkaloid in tobacco leaves,
has been widely studied in the archaeological record due to its apparent stability
and ability to survive over long periods of time
[@eerkensDentalCalculus2018; @raffertyCurrentResearch2012; @tushinghamHuntergathererTobacco2013].

<!--how are compounds incorporated in dental calculus-->
Alkaloids may enter the oral cavity via two pathways: (1) direct
incorporation through oral consumption of alkaloid-containing plants, whether
deliberate or accidental;
and (2) passive diffusion as toxins are transferred from serum to saliva
[@coneInterpretationOral2007],
and then into the oral cavity through the salivary glands. The relation to serum
is why there is often a close correlation between presence (not concentration)
of drugs in oral fluid and blood/serum
[@coneInterpretationOral2007; @willeRelationshipOral2009; @milmanOralFluid2011].
The second pathway allows the
detection of compounds that not only are consumed orally, but also any other mode
of consumption or administration<!--cite e.g. injection or topical-->.
<!-- Additionally, alkaloids will spend shorter time in mouth during direct consumption than
during excretion through saliva, so we expect the second pathway to be more
feasible for detection. -->

Many of the components involved in the formation and growth of dental calculus
originate from oral fluid. Proteins, bacteria, salts and other compounds are
transferred from saliva to biofilms on the tooth surface
[@whiteDentalCalculus1997; @jinSupragingivalCalculus2002].
This may also allow various alkaloids of dietary and medicinal origin to become
incorporated in dental plaque. Dental plaque undergoes frequent mineralisation
events, ultimately causing the entrapped alkaloids and their metabolites to become
preserved within the dental calculus. Barring intentional or accidental removal of
the calculus during life,
burial, excavation, and post-excavation cleaning, the alkaloids can then be detected
by various methods to show a record of consumption during life.

<!--brief description of method validation-->
The method we use in this study was developed in a previous study on dental
calculus from cadavers and validated by comparing the results to compounds detected
in the blood of the same individuals.
All compounds that were detected in the blood were also detected in dental calculus,
with additional compounds present in dental calculus that were not present
in blood, suggesting that dental calculus represents a comprehensive history of
consumption over a long time <!-- rephrase-->[@sorensenDrugsCalculus2021].
<!--summary of results-->
We were able to detect both parent compounds and metabolites, including caffeine,
nicotine, theophylline, and cotinine, in the dental calculus of individuals from
a 19th century Dutch population from Middenbeemster. By detecting these compounds
we are able to show the consumption
of tea and coffee and smoking of tobacco on an individual scale, which is also
confirmed by historic documentation and identification of pipe notches in the
dentition.

### 2023-01-04 inst/paper/_materials.qmd
add: revisions based on co-authors feedback

To reduce the number of factors to account for
in the analysis, we preferentially selected males from the middle adult age category
(35-49). The sample consists of `r nrow(filter(demography, sex == "m"))` males,
`r nrow(filter(demography, sex == "pm"))` probable males,
`r nrow(filter(demography, sex == "pf"))` probable female, and
`r nrow(filter(demography, sex == "f"))` female.
We selected males due to a higher
occurrence of dental calculus deposits than females. Dental calculus was
preferentially selected from incisors.

```{r}
# create tables for demographics if there are tables to spare in the article
demography %>%
  count(sex) %>%
  knitr::kable()

demography %>%
  count(age) %>% 
  knitr::kable()
```

>>>>

To reduce the number of factors to account for
in the analysis, we preferentially selected males from the middle adult age category
(35-49). The sample consists of `r nrow(filter(demography, sex == "m"))` males,
`r nrow(filter(demography, sex == "pm"))` probable males,
`r nrow(filter(demography, sex == "pf"))` probable female, and
`r nrow(filter(demography, sex == "f"))` female.
We selected males due to a higher
occurrence of dental calculus deposits than females (unpublished observation).
Dental calculus was preferentially selected from incisors.

<!-- use plot to show original and replicated sample demographics? -->
```{r}
#| label: fig-sample-demography
#| fig-cap: "Overview of sample demography. Left plot is the first batch and right plot is replication batch. eya = early young adult (18-24); lya = late young adult (25-34); ma = middle adult (35-49); old = old adult (50+)."

# create tables for demographics if there are tables to spare in the article
first_batch <- demography %>%
  #left_join(select(metadata, id, replicated)) %>%
  mutate(sex = factor(sex, levels = c("f", "pf", "pm", "m"))) %>% 
  ggplot(aes(x = age, fill = sex)) +
    geom_bar() +
    scale_fill_viridis_d(drop = FALSE, labels = c("Female", "Female?", "Male?", "Male")) +
    theme_minimal() +
    labs(y = "Count", fill = "Sex") +
    theme(
      panel.grid.major.x = element_blank(),
      axis.title.x = element_blank()
    )

second_batch <- demography %>%
  left_join(select(metadata, id, replicated)) %>%
  filter(replicated == TRUE) %>% 
  mutate(sex = factor(sex, levels = c("f", "pf", "pm", "m"))) %>% 
  ggplot(aes(x = age, fill = sex)) +
    geom_bar() +
    scale_fill_viridis_d(drop = FALSE, labels = c("Female", "Female?", "Male?", "Male")) +
    scale_y_continuous(limits = c(0,25)) +
    theme_minimal() +
    labs(fill = "Sex") +
    theme(
      axis.title = element_blank(),
      axis.text.y = element_blank(),
      axis.ticks.y = element_blank(),
      panel.grid.major.x = element_blank()
    )

first_batch + second_batch + plot_layout(guides = "collect")

grid::grid.draw(grid::textGrob("Age-at-death", x =.45, y=0.01))

# demography %>% 
#   count(sex) %>%
#   knitr::kable()
# 
# demography %>%
#   count(age) %>% 
#   knitr::kable()
```
